Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-13 submission/manuscript.tex
Clean up manuscript and get it to render

\setlength{\droptitle}{-2em}
  \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
Microbiota, and Colorectal Tumors}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

@@

Marc A. Sze\({^1}\), Nicholas A. Lesniak\({^1}\), Mack T. Ruffin
IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)

@@

@@

\linenumbers

\subsection{Abstract}\label{abstract}

\textbf{Background.} Colorectal cancer (CRC) is increasing in prevalence
in individuals under 50 and because of this will be a continuing health
concern for the foreseeable future. The majority of the risk for
developing CRC is attributable to environmental factors. One of these
environmental factors is the microbiota, with certain bacterial
community members being more prevalent in CRC while other taxa are
higher in individuals without tumors. Some of the bacterial species in
taxa that are more abundant in individuals without tumors can use fiber
to produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
model systems. However, the data supporting the importance of SCFAs in
human CRC is less certain. Here, we test the hypothesis that SCFA
concentrations and taxa associated with their production are different
in individuals with colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their fecal concentrations of
acetate, butyrate, and propionate. Analysis also included imputed gene
relative abundance with PICRUSt, metagenomic sequencing on a subset
(n=85) of the total cross-sectional group, and tumor classification and
SCFA prediction models using Random Forest.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
there was also no difference in genes involved with SCFA synthesis
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.70). Finally, there was no
difference between the ability of Random Forest models to predict
patients with adenomas or carcinomas versus individuals without tumors
(P-value \textgreater{} 0.05).

\textbf{Conclusions.} Although our data does not support the hypothesis
that fecal SCFA concentrations are different in the general CRC
population, there still may be specific types of colorectal tumors where
SCFAs may be beneficial for treatment of CRC. Alternatively, our
observations also support the hypothesis that there may be other
metabolites or mechanisms (e.g.~bacterial niche exclusion) that may be
more protective against tumorigenesis and have not been thoroughly
investigated in the context of human CRC.

@@

\subsection{Introduction}\label{introduction}

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US and the prevelance is increasing in
invidiauls under 50 years of age (1, 2). Although there is a genetic
component to the disease, the environment is considered a larger risk
factor for CRC (3). These environmental risk factors include but are not
limited to smoking cigarettes, diet, and the microbiota (4--6). Many of
these environmental risk factors, including the microbiota, are
modifiable. This has lead to the investigation of how the microbiota may
exacerbate or cause tumorigensis (7--9) and whether the bacterial
community is altered in CRC (10, 11). Many of these previous
case/control studies have identified resident bacterial taxa to be
decreased in patients with carcinoma tumors (11--13). Many of the
bacterial species from these resident taxa identified in these
case/control studies actively produce short-chain fatty acids (SCFAs)
from fiber that are part of our general diet (14). The most extensively
studied SCFAs are acetate, butyrate, and propionate (15). These SCFAs
are hypothesized to be the main metabolites involved with protection
against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an
anti-tumorigenic agent. Specific SCFAs have shown positive results
within model systems (16). For example, butyrate has been shown to
inhibit cancer cell growth in \emph{in vitro} systems (17).
Additionally, fiber supplementation in mouse models of CRC caused an
overall reduction in tumor burden while also increasing SCFA
concentrations (18). These exciting results in model systems suggest
that supplementation with food sources that bacteria use to create these
SCFAs may confer beneficial effects against CRC. However, it is
important to note that these model systems provide only preliminary
evidence towards the ability of SCFAs to reduce and treat tumors and the
studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas or individuals
without colon tumors (19). Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (20). These findings argue against
the utility of treatments that aim to use SCFAs to reduce or protect
against tumorigenesis. Given the lack of clear evidence in human studies
of the benefit of SCFAs in CRC, there is a need for more investigation
into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, by
using a seperate cohort it tests previous case/control findings on SCFA
concentrations in individuals with and without tumors. Additionally,
prior investigations grouped patients with adenoma and individuals
without tumors into a single group to compare against patients with
carcinomas. Despite doing this, the suggestion was made that a reduction
in SCFA concentration would also be observed between patients with
adenomas and indivdiuals without colon tumors. We also can test this
suggestion because we have a larger number of patients with adenomas
within our study and do not need to group patients with adenomas and
individuals without colon tumors into a single group. Additionally, we
build upon these observations by assessing the utility of using SCFAs
and Operational Taxonomic Units (OTUs) as a risk stratification tool of
colorectal tumors (adenoma or carcinoma). We also investigate whether
OTUs that are most important to these models are closely associated with
the classification of SCFA concentrations. Collectively, this study
provides important information on the replicability of previous findings
in humans by extensively studying how SCFAs are associated with
colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors. To
provide further support for our SCFA findings we also used PICRUSt (21)
and metagenomic sequencing to investigate differences in genes encoding
enzymes involved with SCFA synthesis between individuals without colon
tumors, patients with adenomas, and patients with carcinomas. Next, we
investigated whether taxa associated with SCFA production were important
to disease classification models. First, using the cross-sectional data,
we analyzed the number of correlations between OTU relative abundance
and SCFA concentrations across individuals without colon tumors and
patients with adenomas or carcinomas. Second, we assessed the affect
adding SCFA concentrations to OTU data had on classification of patients
with adenomas or carcinomas using the Random Forest algorithm (22).
Third, we analyzed how well 16S rRNA gene sequencing predicts SCFA
concentrations. Finally, we compared whether SCFA concentrations replace
important taxa in disease models and whether the same OTUs are the most
important variables used to classify disease and SCFA concentration.
Collectively, this investigation provides additional information as to
whether SCFAs are decreased in patients with colorectal tumors and
provides context as to whether targeting taxa to increase SCFA
concentrations is a viable option to protect against colon
tumorigenesis.

@@

\subsection{Results}\label{results}

\textbf{Decreased SCFA concentrations are not associated with adenoma or
carcinoma tumors.} We used high-performance liquid chromatography (HPLC)
to measure acetate, butyrate, and propionate concentrations of frozen
fecal samples from 490 individuals at a cross-sectional point in time.
There was no difference between individuals without colon tumors (n=172)
and patients with either an adenoma (n=198) or carcinoma (n=120) for any
of the SCFAs measured after multiple comparison correction (P-value
\textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next measured the
concentration of SCFAs in 67 patients with an adenoma (n=41) or
carcinoma (n=26) in which we had pre- and post-treatment fecal samples.
Although there was a general trend for increasing acetate, butyrate, and
propionate concentrations following treatment for tumors, there was no
significant difference between pre- and post-treatment for either
patients with adenomas (P-value \textgreater{} 0.20) or carcinomas
(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

\textbf{Gene abundance for enzymes involved in SCFA synthesis are the
same for individuals without colon tumors and patients with adenomas or
carcinomas.} In order to provide further confirmation and support of our
SCFA concentration results we investigated the genes encoding specific
enzymes involved with SCFA synthesis {[}Table S1{]}. Using this list of
specific genes {[}Table S1{]}, we looked for differences in gene
abundance between individuals without colon tumors and patients with
adenomas or carcinomas. Although we intended to analyze all the genes in
the list {[}Table S1{]}, not all of the KEGG genes were identified
during our analysis. We first analyzed imputed gene relative abundance,
calculated from our OTU data that was generated from 16S rRNA gene
sequencing. We found no difference in any of the imputed gene relative
abundances for enzymes involved with acetate, butyrate, or propionate
synthesis (P-value \textgreater{} 0.90) {[}Table S2{]}. Since butyrate
is one of the most studied SCFAs in the context of CRC, we visualized
the observed lack of difference between indviduals without colon tumors
and patients with adenomas or carcinomas in gene abundance for enzymes
involved with SCFA synthesis using the butyrate kinase gene {[}Figure
2A{]}. Additionally, using a paired Wilcoxon rank-sum test, there also
was no difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and used metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals without colon tumors (n=29) and
patients with adenomas (n=28) or carcinomas (n=28) (P-value
\textgreater{} 0.70) {[}Table S4{]}. This similarity between indviduals
without colon tumors and patients with adenomas or carcinomas is
highlighted again by visualizing the gene abundance for the butyrate
kinase gene {[}Figure 2B{]}. These observations provide evidence that
gene prevalence for enzymes involved in SCFA synthesis does not change
due to colorectal tumors and provides further support for our original
SCFA concentration observations.

\textbf{The number of OTUs positively associated with SCFA concentration
were similar between individuals without colon tumors and patients with
adenomas or carcinomas.} Having found no difference between individuals
without colon tumors and patients with adenomas or carcinomas in SCFA
concentrations or genes encoding enzymes involved with SCFA synthesis,
we next investigated if specific OTUs correlated with SCFA
concentrations. The main goal of this analysis was to identify if there
were OTUs that were significantly associated with SCFA concentrations
and if this was different between individuals that did not have colon
tumors, had an adenoma, or had a carcinoma. To accomplish this we used
Spearman's rho, a non-parametric measure of association, and tested if
there was a correlation that was signficantly greater than zero. We
found that taxa from \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} dominated statistically significant OTU
correlations {[}Figure 3 \& Table S5{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Patients with adenomas also had more positive correlations between OTUs
and SCFA concentrations, but their total number was more similar to
individuals without colon tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. Additionally, the number of positive correlations between OTUs and
SCFA concentrations was similar between individuals without colon tumors
and patients with a carcinoma {[}Figure 3{]}. Finally, when we used
high/low SCFA groups based on the overall median concentration for each
SCFA instead of SCFA concentrations a similar pattern was still observed
{[}Figure S1 \& Table S6{]}. Overall, these results suggest that the
resident taxa that may change the most due to colon tumors may not be
ones that are responsible for the production of acetate, butyrate, or
propionate.

\textbf{SCFA concentrations do not replace important
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
OTUs in Random Forest models built to classify tumors.} Despite the lack
of difference in positive correlations between OTUs and SCFA
concentrations between individuals with and without tumors, OTUs
associated with SCFA concentrations could still be the most important
variables to Random Forest models built to classify patients with
adenomas or carcinomas. We tested this by using the Random Forest
algorithm to build models with OTU abundance data or OTU abundances and
SCFA concentrations to classify normal versus adenoma and normal versus
carcinoma fecal samples. With these models we compared whether OTUs with
taxonomic classification to \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} remained when SCFA
concentrations were added to the model. Additionally, we also compared
whether any of the important OTUs that remained also had significant
correlations with SCFA concentrations. Both our adenoma and carcinoma
models classified patients with a similar degree of success, as measured
by the area under the curve (AUC) (P-value \textgreater{} 0.05)
{[}Figure 4A \& 4D{]}. After the addition of SCFA concentrations to the
adenoma or carcinoma models, many OTUs with taxonomic classification to
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
remained as important variables to the model {[}Figure Figure 4B-C \&
4E-F{]}. After adding SCFA concentrations to the adenoma model, there
were only 2 OTUs significantly associated with SCFA concentration that
were part of the top 10 most important variables and both were postively
associated with acetate, butyrate, or propionate concentrations. When
SCFA concentrations were added to the carcinoma model, only 1 OTU that
was associated with SCFA concentration was in the top 10 most important
variables and it was negatively associated with acetate and butyrate
concentrations. In combination with the previous results on OTU
correlations, these observations provide additional evidence that the
resident taxa that are associated with protection against tumorigeneis
are not ones associated with acetate, butyrate, or propionate
production.

\textbf{The most important OTUs for SCFA concentration and tumor Random
Forest models are not the same.} It is possible that due to the way the
Random Forest algorithm works, OTUs associated with SCFA concentrations
could be downweighted in importance within the adenoma or carcinoma
models when SCFA concentrations are included. To test if this is the
case we used OTU data and built Random Forest models to classify SCFA
concentrations. Overall, the correlation between the predicted and
actual SCFA concentrations were moderately associated with each other
{[}Figure 5A{]}. Additionally, because the training set
R\textsuperscript{2} was always higher than the test set
R\textsuperscript{2}, all SCFA concentration models tended to be over
fit, suggesting that rarer taxa were important for these classifications
{[}Figure 5A{]}. There also was a difference in accuracy based on
whether the fecal sample was from an individual without colon tumors or
from patients with adenomas or carcinomas {[}Figure 5B{]}. When
comparing the adenoma or carcinoma model to the SCFA concentration
models there was minimal overlap between these model's most important
OTUs {[}Figure 4B-C, 4E-F and 5C-E{]}. The only OTU that was in the top
10 most important variables and had overlap between the models was
OTU00167 (\emph{Clostridiales}) {[}Figure 4B-C, 4E-F, 5C-E{]}. Similar
observations were made when using high/low SCFA groups based on the
median SCFA concentration {[}Figure S2{]}. Collectively, these
observations provide evidence that it is possible to identify specific
OTUs associated with higher SCFA concentrations and accordingly these
OTUs belong to taxa known to produce acetate, butyrate, and propionate.
Although it is possible to identify OTUs associated with SCFA
production, overall, our results do not support the hypothesis that SCFA
concentration or OTUs associated with their production are different
between individuals with no colon tumors and patients with adenomas or
carcinomas.

@@

\subsection{Discussion}\label{discussion}

The observations from this study do not support the hypothesis that SCFA
concentrations are different in individuals with tumors. Whether we
directly measured the SCFA concentration or investigated genes that
encoded enzymes used for their production, no difference could be
identified between individuals without tumors and patients with adenomas
or carcinomas {[}Figure 1 \& 2{]}. There were differences in the number
of significant correlations between SCFA concentration and OTU relative
abundance based on whether individuals did not have colon tumors, had an
adenoma, or had a carcinoma. However, SCFA concentrations did not
provide increased model accuracy for tumor classification {[}Figure 3-4
\& S1{]}. In models with SCFA concentrations included, many OTUs that
classified to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} remained as important variables for the model
{[}Figure 4{]}. Additionally, when models using OTU relative abundance
to classify SCFA concentrations were assessed, the OTUs that classified
to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} were not the same as the OTUs that classified to
these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
observations suggest that the resident taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae} that are different
between individuals without colon tumors and patients with adenomas or
carcinomas, are not the same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these
studies have been performed in model systems (16, 17). Many of the
\emph{in vivo} studies use proxies such as fiber supplementation rather
than SCFAs directly (14). Although it is well known that breakdown
products from gut bacteria results in SCFA production, fiber effects on
tumorigenesis may be through other mechanisms in these \emph{in vivo}
model systems. Additionally, the observations in humans on the benefit
of SCFAs in preventing tumorigenesis have been mixed. In previous
case/control studies lower SCFA concentrations were observed in patients
with carcinomas versus those without carcinomas (19). Yet, this is in
contrast to multiple randomized-controlled trials that have found no
difference in tumor recurrence between patients who do and do not get
fiber supplementation (20, 23). In contrast to the \emph{in vivo} model
findings, the observations made in these randomized-controlled trials
would suggest that SCFAs do not prevent or slow tumorigenesis. One
reason for these results is that SCFA concentrations and responses to
fiber vary quite a bit between healthy individuals (24). This
information taken together with our observations would suggest that
either SCFAs provide little to no benefit as an anti-tumorigenic
compound in colorectal cancer or that individuals who do not respond to
fiber supplementation would need to acquire these bacteria to achieve a
benefit.

Another possible alternative explanation as to why no difference in SCFA
concentration between individuals without colon tumors and patients with
adenomas or carcinomas was observed, could be because only certain types
of colorectal cancers are affected by SCFAs. One limitation of current
research into the effect of SCFAs and the microbiota in CRC has been
that all tumors are treated as the same type. However, there are known
differences in the types of mutations that occur (25) and treating all
tumors as equal may actually hide any benefit that could be found in
certain subsets of individuals. Similar to the idea of using
immunotherapy as a targeted treatment option for specific tumors (26),
SCFAs may have beneficial effects for distinct types of colorectal
tumors. Future research will need to test if this is a valid hypothesis.
Regardless of this limitation, our results in combination to previous
randomized controlled trials on fiber supplementation (20, 23) suggests
that using SCFAs as a general treatment for colorectal cancer is
unlikely to provide a reduction in tumorigenesis.

One possible technical limitation is that a fecal sample may not be an
ideal type of bio-specimen and that the effect SCFAs have on
tumorigenesis is only detected in the colon. However, this is unlikely
to be a major confounder. First, most \emph{in vivo} studies, as well as
human studies, have used fecal material in their analysis (18, 19).
Second, previous studies that measure SCFA changes after fiber
supplementation use fecal material to track these responses with a great
deal of success (24). Although there are limitations with the current
research on SCFAs and colorectal tumors, technical limitations are less
likely to be the cause of this. Clearly, our observations provide robust
evidence that these specific metabolites may not be protective or used
as a general treatment option in colorectal cancer. Yet, taxa that are
associated with SCFA production are consistently higher in individuals
without colon tumors than patients with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs, even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Our data is consistent with the literature despite finding no difference
in SCFA concentrations between individuals without colon tumors and
patients with adenomas or carcinomas. In particular, we observed that
the important OTUs in our tumor models were to taxa that are associated
with SCFA production but that these OTUs were not themselves associated
with acetate, butyrate, or propionate production. One of the clearest
results to highlight this within our study were the results showing that
different OTUs from the same taxonomic classification are in tumor and
SCFA Random Forest models {[}Figure 4 \& 5{]}. This leads to the
possibility that protection may be through other routes. For example,
there could be a different pathway or other less extensively studied
metabolites that provide the necessary protection against tumorigenesis.
Alternatively, previous research has shown that non-resident oral
microbes, such as \emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}, and \emph{Peptostreptococcus} are associated with
carcinomas (6, 12, 13) and can increase inflammation (28). When these
non-resident oral microbes increase inflammation they change the
existing environment to one in which is more hospitable for themselves
(28). It is possible that these resident microbes provide protection not
through a metabolite but instead through active exclusion of these
non-resident oral microbes which in turn prevents them from establishing
themselves within the community. Overall, even though we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that our results open up many new and exciting avenues for
investigation into how metabolites and the microbiota can prevent
tumorigenesis.

@@

\subsection{Conclusions}\label{conclusions}

Our observations found no difference in SCFA concentration or in genes
encoding enzymes involved in their production between individuals
without colon tumors and patients with either adenoma or carcinoma
tumors. The SCFA concentration also did not improve classification of
adenoma or carcinoma tumors. The most important OTUs to these models
classifying patients with adenomas or carcinomas were mostly not
associated with SCFA production. Although these results are different
than other reports in the literature, they do align with the randomized
controlled trials that have tested fiber use in preventing colorectal
tumor recurrence. Additionally, these observations suggest that resident
microbes that are not involved in SCFA production may be the important
resident community members involved with preventing tumorigenesis. By
focusing on alternative mechanisms that are associated with these
non-SCFA producing resident microbes or SCFA effects on distinct
colorectal cancer groups the identification of more promising
therapeutic options for use in treating colorectal cancer may be found.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (29, 30). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment of a patient's tumor
(adenoma n =41 and carcinoma n = 26). For patients undergoing treatment
for their tumor the length of time between their initial and follow up
sample ranged from 188 - 546 days. Our use of treatment has been
previously defined as encompassing removal of a tumor (surgery or
colonoscopy) with or without chemotherapy and radiation (29). Diagnosis
of tumor was made by colonoscopic examination and histopathological
review of biopsies obtained (29, 30). The University of Michigan
Institutional Review Board approved the study and informed consent was
obtained from all participants in accordance to the guidelines set out
by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (24). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (24) and Figure 1 here).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptiClust algorithm (35).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (21) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=490) containing a total of 85
individuals (normal n=29, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (36). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (37). Open Reading Frames (ORFs) were
identified using Prodigal (38), counts generated using Diamond (39),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (40), and OPF alignment used the KEGG database (41).

>>>>

\setlength{\droptitle}{-2em}

  \title{Revisiting the Relationship between Short-Chain Fatty Acids, the
Microbiota, and Colorectal Tumors}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

@@

Marc A. Sze\({^1}\), Begüm Topçuoğlu\({^1}\), Nicholas A.
Lesniak\({^1}\), Mack T. Ruffin IV\({^2}\), Patrick D.
Schloss\({^1}\)\({^\dagger}\)

@@

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{introduction-400-words}{%
\subsection{Introduction (400 words)}\label{introduction-400-words}}

Colorectal cancer is the third leading cancer-related cause of death
within the United States (1, 2). Less than 10\% of cases can be
attributed to genetic risk factors (3). This leaves a significant role
for environmental, behavioral, and other factors such as smoking and
diet (4, 5). Colorectal cancer is thought to be initiated by a series of
mutations that accumulate as mutated cells proliferate leading to
adenomatous lesions, which are succeeded by carcinomas (3). Throughout
this progression, there are ample opportunities for bacterial
populations to create mutations, induce inflammation, and accelerate
tumorigenesis (6). Numerous studies in murine models have supported this
model (7--11). Additional cross sectional studies in humans have
identified microbiome-based biomarkers of disease (12--20). These
studies suggest that in some cases, it is the loss of bacterial
populations that produce short-chain fatty acids (SCFAs) that results in
increased inflammation and tumorigenesis.

SCFAs have have anti-inflammatory and anti-proliferative activities
(21--24). Furthermore, manipulation of SCFAs in mouse models of
colorectal cancer by direct supplementation or feeding of fiber caused
an overall reduction in tumor burden (25, 26). These results suggest
that supplementation with substrates that bacteria can ferment to
produce SCFAs may confer beneficial effects against colorectal cancer.
Regardless, there is a lack of evidence that increasing SCFA
concentrations can protect against colorectal cancer in humans. Based on
similar observations, many microbiome studies use the concentrations of
SCFAs and the presence of 16S rRNA gene sequences from organisms and the
genes involved in producing them as a biomarker of a healthy microbiota
(27--30). Case-control studies that have investigated SCFA
concentrations in colorectal cancer found that patients with carcinomas
had lower concentrations of SCFAs versus patients with adenomas or
individuals without colon tumors (31). Although this would argue that
increasing SCFA concentrations could be protective against
tumorigenesis, in randomized controlled trials fiber supplementation has
been inconsistently associated with protection against tumor formation
and recurrence (32--36). These findings temper enthusiasm for treatments
that aim to use SCFAs as biomarkers or protection against tumorigenesis.

To better understand the connection between colorectal cancer, the
microbiome, and SCFAs, we quantified the concentration of SCFAs in feces
of previously characterized individuals with normal colons, adenomas,
and carcinomas and subset of those individuals after they underwent
treatment for their lesions (14, 16). Our goals included: (1) testing
whether there was an association between SCFA concentration and tumor
status, (2) determining whether SCFA concentrations could be used as
biomarkers to improve the detection of colonic lesions, and (3)
predicting SCFA concentrations based on the relative abundance of fecal
bacteria and their genes.

@@

\hypertarget{results-700-words}{%
\subsection{Results (700 words)}\label{results-700-words}}

\begin{itemize}
\tightlist
\item
  There is no association between presence of tumors and SCFA
  concentration

  \begin{itemize}
  \tightlist
  \item
    Cross sectional
  \item
    Aggregating SCFAs does not improve association with CRC
  \item
    Lack of change pre and post treatment
  \item
    Aggregating SCFAs does not improve association with CRC
  \end{itemize}
\item
  Adding SCFA data to diagnostic models does not improve performance

  \begin{itemize}
  \tightlist
  \item
    Combining 16S and SCFA does not improve prediction of CRC
  \item
    OTU, genus, OPF, kegg
  \end{itemize}
\item
  Community structure is not predictive of SCFA production

  \begin{itemize}
  \tightlist
  \item
    Random forest regression
  \item
    16S
  \item
    picrust
  \item
    shotgun metagenomic data
  \end{itemize}
\end{itemize}

@@

\hypertarget{discussion-200-words}{%
\subsection{Discussion (200 words)}\label{discussion-200-words}}

\begin{itemize}
\tightlist
\item
  Unlikely that SCFAs are the primary mechanism that limits
  tumorigenesis
\item
  Limitations

  \begin{itemize}
  \tightlist
  \item
    Observing the community once the tumor is developed
  \item
    Not looking at mucosal communities
  \end{itemize}
\item
  Need to identify other possible mechanisms that drive and prevent
  tumorigenesis
\item
  Supports evidence that fiber consumption has a null effect on
  tumorigenesis
\end{itemize}

@@

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (14, 37). In brief, this study used fecal
samples obtained at either a single cross-sectional time point
(n=\textbf{r
length(metai\(sample)**) or from before (pre-) and after (post-) treatment of a patient's tumor (adenoma n =**r length(filter(metaf, dx == "adenoma") %>% select(initial) %>% pull())** and carcinoma n = ** length(filter(metaf, dx == "cancer") %>% select(initial) %>% pull())**). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from **r min(metaf
\)time)} - \textbf{r max(metaf\$time)} days. Our use of treatment has
been previously defined as encompassing removal of a tumor (surgery or
colonoscopy) with or without chemotherapy and radiation (37). Diagnosis
of tumor was made by colonoscopic examination and histopathological
review of biopsies obtained (14, 37). The University of Michigan
Institutional Review Board approved the study and informed consent was
obtained from all participants in accordance to the guidelines set out
by the Helsinki Declaration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol that used a High-Performance Liquid Chromatography (HPLC)
machine (38). The following changes to this protocol included the use of
frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the actual
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These methodological changes did not
affect the overall median concentrations of these SCFAs between the two
studies (see Table 1 (38) and Figure 1 here).

\textbf{16S rRNA gene sequencing.} The workflow and processing have been
previously described (37, 39, 40). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(41). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (42). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptiClust algorithm (43).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (44) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

\textbf{Obtaining Operational Protein Families from metagenomes.} A
subset of the cross-sectional group (n=\textbf{r
length(metai\(sample)**) containing a total of **r length(geof_samples\)X1)}
individuals (normal n=\textbf{r filter(geof\_samples, X30 ==
``Healthy'') \%\textgreater{}\% pull(X30) \%\textgreater{}\% length()},
adenoma n=\textbf{r filter(geof\_samples, X30 == ``Adenoma'')
\%\textgreater{}\% pull(X30) \%\textgreater{}\% length()}, and carcinoma
n=\textbf{r filter(geof\_samples, X30 == ``Cancer'') \%\textgreater{}\%
pull(X30) \%\textgreater{}\% length()}) was shotgun sequenced on an
Illumina HiSeq using 125 bp paired end reads and a previously described
method (16). Briefly, the sequences were quality filtered and sequences
aligning to the human genome were removed prior to contig assembly with
MEGAHIT (45). Open Reading Frames (ORFs) were identified using Prodigal
(46), counts generated using Diamond (47), subsequent clustering into
Operational Protein Families (OPFs) used mmseq2 (48), and OPF alignment
used the KEGG database (49).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (8,
49) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification and regression models via the
caret package (50, 51). This was repeated on 100 different 80/20 splits
of the data to generate a reasonable range for the AUC of the model. The
reported AUCs, unless otherwise specified, are for the test sets. The
classification models were built to group normal versus adenoma, normal
versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (52). Generally, a Kruskal-Wallis rank sum
test with a Dunn's post-hoc test was used to assess differences between
individuals without colon tumors, patients with adenomas, and patients
with carcinomas. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons {[}\textbf{XXXXXX}{]}. First, we
assessed differences in SCFA concentrations measured by HPLC between
individuals with normal colons and patients with tumors (adenoma or
carcinoma). We then analyzed whether SCFA concentrations changed in
patients with an adenoma or carcinoma pre- versus post-treatment. Next,
the imputed gene counts of genes encoding enzymes involved in SCFA
synthesis was tested. Additionally, metagenomic sequencing counts for
important genes involved with SCFA production were analyzed. From here
we analyzed the number of significant positive and negative correlations
between OTU relative abundance and SCFA concentrations in individuals
without colon tumors and patients with adenomas or carcinomas using
Spearman's rho. Next, we assessed whether OTUs alone or OTUs and SCFAs
were better able to classify individuals with and without tumors using
Random Forest models. Finally, models to classify the actual SCFA
concentration or high/low SCFA concentration based on the median of each
SCFA using 16S rRNA gene sequencing data was created using the Random
Forest algorithm. For all Random Forest models, the assessment of the
most important variables was based on the top 10 features (OTUs or
SCFAs) using the mean decrease in accuracy.

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}
